---
title: "Clustering Analysis"
subtitle: "Master in Bioinformatics, University of Murcia"
author: "Joaquín Cantón Sandoval (joaquin.canton@um.es)"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    highlight: kate
    number_sections: true
    theme: spacelab
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

```{r}
library(pvclust)
library(dplyr)
library(ggplot2)
library(FactoMineR)
library(factoextra)
library(tidyr)
```

```{r}
datos_filtrados = read.csv("../datos_filtrados.csv")

# Find parameters related to tracks
track_parameters <- grep("Track", names(datos_filtrados), value = TRUE)

to_remove <- c("TrackID", "TrackID_sd", "Track.Position.Z.Start",
              "Track.Position.Z.Mean",
                "Track.Position.Y.Start",
                "Track.Position.Y.Mean",
                "Track.Position.X.Start",
                "Track.Position.X.Mean",
               "TrackImage.CH2",
               "TrackImage.CH1", "TrackImage", "Track.Center.of.Image.Mass.Z.CH2", "Track.Center.of.Image.Mass.Z.CH1",
               "Track.Center.of.Image.Mass.Y.CH1", "Track.Center.of.Image.Mass.Y.CH2", "Track.Center.of.Image.Mass.X.CH1",
               "Track.Center.of.Image.Mass.X.CH2", "Track.Center.of.Homogeneous.Mass.Z", "Track.Center.of.Homogeneous.Mass.Y",
               "Track.Center.of.Homogeneous.Mass.X")

# Actualizar la lista de parámetros de track eliminando los no deseados
track_parameters <- setdiff(track_parameters, to_remove)

# Ver la lista actualizada
head(track_parameters)

boxplot(datos_filtrados[track_parameters])
scaled_filtrados = as.data.frame(scale(datos_filtrados[track_parameters]))
boxplot(scaled_filtrados)

# Delete columns with NA
scaled_filtrados <- scaled_filtrados %>%
  dplyr::select(-c(Track.Number.of.Branches, Track.Number.of.Fusions, Track.Number.Of.Generations, Time.Since.Track.Start))

# Check the dimensions of the updated dataframe
print(dim(scaled_filtrados))
```

# Clustering


## Distance calculation

We begin by calculating the distance matrix on scaled data, which will serve as the basis for our hierarchical clustering.

```{r}
compared.dist = dist(scaled_filtrados)
```

We then apply various hierarchical clustering methods to this distance matrix.

```{r}
complete <- hclust(compared.dist, method = "complete")
avg <- hclust(compared.dist, method = "average")
single <- hclust(compared.dist, method = "single")
centroid <- hclust(compared.dist, method = "centroid")
ward <- hclust(compared.dist, method = "ward.D")
```

Next, we compute the cophenetic correlation coefficient for each method to evaluate their performance.

```{r}
complete <- cor(compared.dist, cophenetic(complete))
average <- cor(compared.dist, cophenetic(avg))
single <- cor(compared.dist, cophenetic(single))
centroid <- cor(compared.dist, cophenetic(centroid))
ward <- cor(compared.dist, cophenetic(ward))

mejor <- data.frame(complete, single, centroid, average, ward)
round(mejor, 4)
```

## Clustering

### Euclidean distance and Average method

```{r, eval=FALSE}
# Perform clustering analysis with pvclust
set.seed(123)  
pv_result_track_average <- pvclust(t(scaled_filtrados), method.dist="euclidean", method.hclust="average", nboot=1000)

saveRDS(pv_result_track_average, "pv_result_ave.RDS")
```

```{r}
pv_result_track_average = readRDS("../pv_result_ave.RDS")
# Visualize the result with p values of AU and BP
plot(pv_result_track_average)
```

### Euclidean distance and Complete method

```{r, eval=FALSE}
# Perform clustering analysis with pvclust
set.seed(123)  
pv_result_track_complete <- pvclust(t(scaled_filtrados), method.dist="euclidean", method.hclust="complete", nboot=1000)
saveRDS(pv_result_track_complete, "pv_result_complete.RDS")
```

```{r}
pv_result_track_complete = readRDS("../pv_result_complete.RDS")
# Visualize the result with p values of AU and BP
plot(pv_result_track_complete)
```

### Correlation distance and Average method

```{r, eval=FALSE}
# Perform clustering analysis with pvclust
set.seed(123)  # Para reproducibilidad
pv_result_track_average_cor <- pvclust(t(scaled_filtrados), method.dist="correlation", method.hclust="average", nboot=1000)
saveRDS(pv_result_track_average_cor, "pv_result_ave_cor.RDS")
```

```{r}
pv_result_track_average_cor = readRDS("../pv_result_ave_cor.RDS")
# Visualize the result with p values of AU and BP
plot(pv_result_track_average_cor)
```


## Represent PCA colored by cluster

```{r}
# Extracting the hclust object from pvclust results
hclust_obj = pv_result_track_average_cor$hclust

# Cutting the dendrogram to create two clusters
clusters = cutree(hclust_obj, k = 2)  # k is the number of clusters
# Ensure that the order of rows in original_data matches the one used in pvclust
scaled_filtrados$Cluster = clusters

# Perform PCA
filtrados_track_pca <- PCA(scaled_filtrados, ncp = 58, graph = FALSE)

filtrados_pca_data <- data.frame(filtrados_track_pca$ind$coord)
filtrados_pca_data$Cluster = as.factor(clusters)  # Ensure clusters are factor type for coloring

# Plot PCA with clusters as the coloring variable
fviz_pca_ind(filtrados_track_pca, 
             label = "none",  # No labels on points
             habillage = filtrados_pca_data$Cluster,  # Color by cluster
             addEllipses = TRUE,  # Add confidence ellipses
             ellipse.level = 0.95,  # 95% confidence level
             palette = c("#00AFBB", "#E7B800", "#FC4E07"),  # Colors, adjust as needed
             geom = "point") +  # Use points for individuals
  ggtitle("PCA of Filtered Neutrophils by Cluster") +  # Update title to reflect clustering
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),  # Center title
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis text for better readability
    axis.title.x = element_text(face = "bold"),  # Bold x-axis title
    axis.title.y = element_text(face = "bold")  # Bold y-axis title
  )
```

