---
title: "Multivariate Analysis"
subtitle: "Master in Bioinformatics, University of Murcia"
author: "Joaquín Cantón Sandoval (joaquin.canton@um.es)"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    highlight: kate
    number_sections: true
    theme: spacelab
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

```{r}
library(stats)
library(knitr)
```


```{r}
neutrophils = readRDS("neutrofilos_unique.RDS")
```

## Filter track parameters

```{r}
# Find parameters related to tracks
track_parameters <- grep("Track", names(neutrophils), value = TRUE)

to_remove <- c("TrackID", "TrackID_sd", "Track.Position.Z.Start",
              "Track.Position.Z.Mean",
                "Track.Position.Y.Start",
                "Track.Position.Y.Mean",
                "Track.Position.X.Start",
                "Track.Position.X.Mean",
               "TrackImage.CH2",
               "TrackImage.CH1", "TrackImage", "Track.Center.of.Image.Mass.Z.CH2", "Track.Center.of.Image.Mass.Z.CH1",
               "Track.Center.of.Image.Mass.Y.CH1", "Track.Center.of.Image.Mass.Y.CH2", "Track.Center.of.Image.Mass.X.CH1",
               "Track.Center.of.Image.Mass.X.CH2", "Track.Center.of.Homogeneous.Mass.Z", "Track.Center.of.Homogeneous.Mass.Y",
               "Track.Center.of.Homogeneous.Mass.X")

# Actualizar la lista de parámetros de track eliminando los no deseados
track_parameters <- setdiff(track_parameters, to_remove)

# Ver la lista actualizada
head(track_parameters)
```

# Multivariate Analysis

After identifying significant effects in univariate tests, it's essential to adjust for related factors through multivariate analysis, especially given the complex interactions in biological and experimental contexts. This approach corrects p-values for multiple comparisons and confounding factors, enhancing the precision and biological relevance of statistical results.

## Analysis Based on Condition Adjusting for Individual Variability

Here, we focus on how treatment conditions affect a specific variable—mean track speed (Track.Speed.Mean)—while adjusting for individual variability represented by the factor pID.x.

```{r}
modelo <- glm(Track.Speed.Mean ~ Condition + factor(pID.x), 
              family = gaussian(), data = neutrophils)

# Display model summary to view results
summary(modelo)

```

The fitted model indicates significant effects of the treated condition on neutrophil track speed. Specifically, the coefficient for Conditiontreated is significantly negative, suggesting that the treated condition reduces the mean track speed, adjusting for individual differences.

## Model Dispersion and Deviance Analysis

```{r}
# Initialize a dataframe to store significant results
significant_results <- data.frame(Parameter = character(),
                                  P_Value = character(),
                                  Significance = character(),
                                  stringsAsFactors = FALSE)

# Loop to fit a GLM for each track parameter
for (param in track_parameters) {
  model <- glm(formula = as.formula(paste(param, "~ Condition + factor(pID.x)")),
               family = gaussian(), data = neutrophils)
  summary_model <- summary(model)
  interaction_p_value <- coef(summary_model)["Conditiontreated", "Pr(>|t|)"]

  if (!is.na(interaction_p_value) && interaction_p_value < 0.05) {
    # Determine significance asterisks
    significance <- if(interaction_p_value < 0.0001) "****"
                    else if(interaction_p_value < 0.001) "***"
                    else if(interaction_p_value < 0.01) "**"
                    else if(interaction_p_value < 0.05) "*"
                    else ""
    
    significant_results <- rbind(significant_results, data.frame(Parameter = param,
                                                                 P_Value = formatC(interaction_p_value, format = "e", digits = 2),
                                                                 Significance = significance,
                                                                 stringsAsFactors = FALSE))
  }
}
# Optionally, sort the dataframe by p-value to see the most significant results at the top
significant_results <- significant_results[order(significant_results$P_Value), ]
```

### Data display

```{r}
# Use kable to display the table more elegantly
kable(significant_results, format = "html", caption = "GLM Results for Various Parameters Comparing Condition and pID.x")

```

