---
title: "PCA Analysis"
subtitle: "Master in Bioinformatics, University of Murcia"
author: "Joaquín Cantón Sandoval (joaquin.canton@um.es)"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    highlight: kate
    number_sections: true
    theme: spacelab
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

```{r}
library(FactoMineR)
library(factoextra)
library(ggplot2)
library(viridis)
library(viridisLite)
library(dplyr)
library(tidyr)
```


# Principal Component Analysis (PCA)
This section of the analysis involves conducting a Principal Component Analysis (PCA) on neutrophil track parameters. The goal is to reduce the dimensionality of our data, enabling us to observe patterns and trends that are not readily visible in high-dimensional data. This approach helps us identify the main sources of variation among neutrophils under different experimental conditions, such as their interaction with keratinocytes or other neutrophils, as well as under different treatment conditions. We will visualize these patterns through graphs that highlight the differences and similarities in data distribution, thus facilitating a clearer and more direct interpretation of the complexities inherent in our biological data.


## Data loading 

```{r}
neutrophils = readRDS("neutrofilos_unique.RDS")
```

## Filter track parameters

```{r}
# Find parameters related to tracks
track_parameters <- grep("Track", names(neutrophils), value = TRUE)

to_remove <- c("TrackID", "TrackID_sd", "Track.Position.Z.Start",
              "Track.Position.Z.Mean",
                "Track.Position.Y.Start",
                "Track.Position.Y.Mean",
                "Track.Position.X.Start",
                "Track.Position.X.Mean",
               "TrackImage.CH2",
               "TrackImage.CH1", "TrackImage", "Track.Center.of.Image.Mass.Z.CH2", "Track.Center.of.Image.Mass.Z.CH1",
               "Track.Center.of.Image.Mass.Y.CH1", "Track.Center.of.Image.Mass.Y.CH2", "Track.Center.of.Image.Mass.X.CH1",
               "Track.Center.of.Image.Mass.X.CH2", "Track.Center.of.Homogeneous.Mass.Z", "Track.Center.of.Homogeneous.Mass.Y",
               "Track.Center.of.Homogeneous.Mass.X")

# Actualizar la lista de parámetros de track eliminando los no deseados
track_parameters <- setdiff(track_parameters, to_remove)

# Ver la lista actualizada
head(track_parameters)
```

## Data scaling

```{r}
# Plotting original data
boxplot(neutrophils[, track_parameters])

# Scaling data
track.scaled <- as.data.frame(scale(neutrophils[, track_parameters]))

# Plotting scaled data
boxplot(track.scaled[, track_parameters])
```

### Remove columns with NAs

```{r}
# Removing columns with NA
track.scaled <- track.scaled %>%
  dplyr::select(
    -Track.Number.of.Branches, 
    -Track.Number.of.Fusions, 
    -Track.Number.Of.Generations, 
    -Time.Since.Track.Start)
```

## PCA

```{r}
# Performing PCA on your numerical data
neutrofilos_pca <- PCA(track.scaled, ncp = 58, graph = TRUE)
```

### Visualize PCA

```{r}
# Visualize only the points of PCA
fviz_pca_ind(neutrofilos_pca, 
             label = "none",  # Ensure no point labels are shown
             habillage = neutrophils$Condition,  # Color by migration type
             addEllipses = TRUE,  # Add confidence ellipses
             ellipse.level = 0.95)  # Confidence level for ellipses
  ggtitle("PCA of neutrophils by condition")  # Add plot title
  theme_minimal()  # Apply a minimalistic theme
```

## Get the ideal number of components

```{r}
# Obtain the variance explained by each component
var_explicada <- neutrofilos_pca$eig[, "percentage of variance"]

# Calculate the cumulative explained variance
var_acumulada <- cumsum(var_explicada)

# Find the number of components that explain 80% of the variance
num_componentes_80 <- which.max(var_acumulada >= 80)
num_componentes_80
```

```{r}
# Crear el barplot con colores condicionales
barplot(var_acumulada, names.arg = paste0("Comp.", 1:length(var_acumulada)),
        xlab = "Componentes principales", ylab = "Varianza explicada acumulada",
        col = ifelse(var_acumulada >= 80, "lightgray", "yellow"),
        main = "Varianza explicada acumulada por componentes principales")

# Añadir línea horizontal roja en el 80% de varianza explicada acumulada
abline(h = 80, col = "red", lty = 2)

# Añadir línea vertical azul en la componente 31
abline(v = 17, col = "blue", lwd = 2)  # `lwd` ajusta el grosor de la línea
```

## Get variable loading matrix

```{r}
# Get variable loading matrix
loadings <- get_pca_var(neutrofilos_pca)$coord


# Select up to Dimension 5
loadings_subset <- loadings[, 1:5]

# Sort columns from Dim1 to Dim5
loadings_subset <- loadings_subset[, order(colnames(loadings_subset))]

# Plot the heatmap
heatmap(loadings_subset, col = viridis(20))
```

## Function to obtain the most relevant genes by dimension

```{r}
# Define a function to obtain the most relevant genes by dimension
get_top_genes <- function(loadings_matrix, dimension, n = 10) {
  # Get the loads for the specified dimension
  loadings_dim <- loadings_matrix[, dimension]
  
  # Sort the genes by their loadings in the dimension
  sorted_genes <- sort(abs(loadings_dim), decreasing = TRUE)
  
  # Get the most positive and negative genes
  top_positive <- names(sorted_genes)[head(which(loadings_dim >= 0), n)]
  top_negative <- names(sorted_genes)[head(which(loadings_dim < 0), n)]
  
  return(list(positive = top_positive, negative = top_negative))
}

# Get the most relevant genes for each dimension (for example, for the first 5 dimensions)
top_genes_per_dimension <- lapply(1:5, function(dim) get_top_genes(loadings, dim))

# Show results
for (dim in 1:5) {
  cat("Dim.", dim, ":\n")
  cat("Positive:", paste(top_genes_per_dimension[[dim]]$positive, collapse = ", "), "\n")
  cat("Negative:", paste(top_genes_per_dimension[[dim]]$negative, collapse = ", "), "\n\n")
}
```

# Filter data from PCA

```{r}
# Coordinates of individuals in the PCA
coords <- neutrofilos_pca$ind$coord

# Extract points of interest from the 'treated' group
treated_points <- coords[neutrophils$Condition == "treated", ]

# Find the point with the maximum value on PC1
max_pc1 <- treated_points[which.max(treated_points[, "Dim.1"]), ]

# Find the sixth minimum value in PC2
# Sort PC2 in ascending order and select the sixth
sixth_min_pc2_index <- order(treated_points[, "Dim.2"], decreasing = FALSE)[6]  # Take the sixth smallest index
sixth_min_pc2 <- treated_points[sixth_min_pc2_index, ]

# Calculate the slope of the line joining these two points
slope <- (sixth_min_pc2["Dim.2"] - max_pc1["Dim.2"]) / (sixth_min_pc2["Dim.1"] - max_pc1["Dim.1"])

# Define the intersection
intercept <- max_pc1["Dim.2"] - slope * max_pc1["Dim.1"]

# Extend the line to the plot boundaries
x_limits <- range(coords[, "Dim.1"])  # Determine the limits of Dim.1
y_limits <- intercept + slope * x_limits  # Calculate the new limits of y based on the slope

# View only PCA points
plot <- fviz_pca_ind(neutrofilos_pca, 
             label = "none",  # Ensures no point labels are displayed
             habillage = neutrophils$Condition,  # Colored by condition
             addEllipses = TRUE,  # Add confidence ellipses
             ellipse.level = 0.95) + # Confidence level for ellipses
  geom_segment(aes(x = x_limits[1], y = y_limits[1], 
                   xend = x_limits[2], yend = y_limits[2]),
               linetype = "dashed", color = "blue", size = 1) +
  ggtitle("PCA of neutrophils by condition") +  
  theme_minimal() +  
  theme(
    plot.title = element_text(hjust = 0.5),  
    axis.text.x = element_text(angle = 45, hjust = 1),  
    axis.title.x = element_text(face = "bold"),  
    axis.title.y = element_text(face = "bold")  
  )

# Print the plot
print(plot)
```

## Label neutrophils as "Potentially Benefitial" and "Potentially Dangerous"

```{r}
# Add the data column to the coordinate dataframe
coords_with_data <- cbind(coords, neutrophils)

# Filter data based on position relative to the diagonal line
datos_filtrados <- coords_with_data %>%
  filter(Dim.2 < (slope * Dim.1 + intercept))  # Points below the line
datos_filtrados$Group <- "Potentially Dangerous"

resto <- coords_with_data %>%
  filter(Dim.2 >= (slope * Dim.1 + intercept))  # Points above the line
resto$Group <- "Potentially Beneficial"

# Combine datasets for visualization
combined_data <- rbind(datos_filtrados, resto)

# View results to ensure correct separation
ggplot(combined_data, aes(x = Dim.1, y = Dim.2, color = Group)) +
  geom_point() +
  geom_segment(aes(x = min(coords[, "Dim.1"]), y = intercept + slope * min(coords[, "Dim.1"]),
                   xend = max(coords[, "Dim.1"]), yend = intercept + slope * max(coords[, "Dim.1"])),
               linetype = "dotted", color = "black", size = 1.5) +
  scale_color_manual(values = c("Potentially Dangerous" = "red", "Potentially Beneficial" = "blue")) +
  ggtitle("PCA with Diagonal Separation") +
  theme_minimal() +
  labs(color = "Grupo")  
```



