---
title: "Random Forest for Track Parameters"
subtitle: "Master in Bioinformatics, University of Murcia"
author: "Joaquín Cantón Sandoval (joaquin.canton@um.es)"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    highlight: kate
    number_sections: true
    theme: spacelab
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

# Introduction

This report presents a detailed analysis using Machine Learning techniques to study track parameters in neutrophils. The aim is to optimize and predict treatment conditions using the Random Forest algorithm and explore different configurations of this model.

```{r}
# Load necessary libraries
library(readr)        # For data loading
library(ggplot2)      # For data visualization
library(caret)        # For Machine Learning modeling
library(cluster)      # For k-means clustering
library(factoextra)   # For additional clustering visualizations
library(randomForest) # For Random Forest
library(fastDummies)  # For dummy calculation
library(dplyr)
library(ggbeeswarm)
library(tidyverse)
```

```{r}
# Load the data
neutrofilos_unique = readRDS("../neutrofilos_unique.RDS")

set.seed(1234)
# Identify factor columns
factor_parameters <- names(neutrofilos_unique)[sapply(neutrofilos_unique, is.factor)]
track_parameters <- grep("Track", names(neutrofilos_unique), value = TRUE)

to_remove <- c("TrackID", "TrackID_sd", "Track.Position.Z.Start",
               "Track.Position.Z.Mean", "Track.Position.Y.Start", 
               "Track.Position.Y.Mean", "Track.Position.X.Start", 
               "Track.Position.X.Mean", "TrackImage.CH2", "TrackImage.CH1", 
               "TrackImage", "Track.Ar1.Z", "Track.Ar1.Y", "Track.Ar1.X", 
               "Track.Image", "Track.Ar1.Mean", "pID.x", "Time.Since.Track.Start", 
               "Track.Number.of.Fusions", "Type.Of.Migration", "Track.Number.Of.Generations", 
               "Track.Number.of.Branches", "Track.Center.of.Image.Mass.Z.CH2", "Track.Center.of.Image.Mass.Z.CH1",
               "Track.Center.of.Image.Mass.Y.CH1", "Track.Center.of.Image.Mass.Y.CH2", "Track.Center.of.Image.Mass.X.CH1",
               "Track.Center.of.Image.Mass.X.CH2", "Track.Center.of.Homogeneous.Mass.Z", "Track.Center.of.Homogeneous.Mass.Y",
               "Track.Center.of.Homogeneous.Mass.X")

# Combine track parameters with factorial parameters
track_parameters <- unique(c(track_parameters, factor_parameters))
# Update the list of track parameters by removing unwanted ones
track_parameters <- setdiff(track_parameters, to_remove)


neutrofilos_unique = neutrofilos_unique[track_parameters]
```

# Data Preparation and Split

Description of data preparation and its division into training and testing sets.

```{r}
# Data preparation
neutrofilos_unique$Condition <- as.factor(neutrofilos_unique$Condition)
set.seed(123)
trainIndex <- createDataPartition(neutrofilos_unique$Condition, p = .8, list = FALSE)
neutrofilosTrain <- neutrofilos_unique[trainIndex, ]
neutrofilosTest <- neutrofilos_unique[-trainIndex, ]
```

# Training the Random Forest Model

Description of the Random Forest model training process.

```{bash, eval=FALSE}
# Training control setup
fitControl <- trainControl(
  method = "cv",
  number = 10,
  savePredictions = "final"
)

# Model training
rf_model <- train(
  Condition ~ ., data = neutrofilosTrain,
  method = "rf",
  trControl = fitControl,
  ntree = 500
)
```

# Script for the server

```{bash, eval=FALSE}
#!/bin/bash

#SBATCH --job-name=neutrofilos_rf_training    # Nombre del trabajo
#SBATCH --output=rf_%u.%x.%j.out        # Salida estandar (stdout)
#SBATCH --error=rf_%u.%x.%j.err         # Salida de error (stderr)
#SBATCH --cpus-per-task=4       #
#SBATCH --chdir=/home/alumno04/TFM

module load R                     
Rscript /home/alumno04/TFM/rf_neutrofilos.R  
```

# Optimizing Random Forest

```{r}
rf.cv.neutrofilos = readRDS("../randomforest_neutrofilos.RDS")

# Create a data frame with the updated results
results <- data.frame(
  ntrees = c(rep(500, 7), rep(1500, 7), rep(2500, 7)),
  mtry = rep(2:8, 3),
  Accuracy = c(
    0.8842510, 0.8905258, 0.9014633, 0.8998760, 0.9030258, 0.9014881, 0.9045883,  # 500 trees
    0.8763145, 0.8919395, 0.8982143, 0.8966270, 0.8997520, 0.9013145, 0.8981895,  # 1500 trees
    0.8732887, 0.8920387, 0.8982887, 0.9061012, 0.9092262, 0.9061012, 0.9107887   # 2500 trees
  ),
  Kappa = c(
    0.6269300, 0.6595833, 0.6978788, 0.6989808, 0.7074392, 0.7063465, 0.7142408,  # 500 trees
    0.5864067, 0.6560980, 0.6811084, 0.6799673, 0.6938638, 0.6957510, 0.6916841,  # 1500 trees
    0.5831214, 0.6578878, 0.6872619, 0.7156072, 0.7276882, 0.7165469, 0.7314174   # 2500 trees
  )
)

# Plot Accuracy as a function of mtry for each ntrees
library(ggplot2)

ggplot(results, aes(x = mtry, y = Accuracy, group = ntrees, colour = as.factor(ntrees))) +
  geom_line() +
  geom_point() +
  labs(title = "Accuracy as a Function of mtry and ntrees",
       x = "Mtry",
       y = "Accuracy",
       colour = "Ntrees") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

# Plot Kappa as a function of mtry for each ntrees
ggplot(results, aes(x = mtry, y = Kappa, group = ntrees, colour = as.factor(ntrees))) +
  geom_line() +
  geom_point() +
  labs(title = "Kappa as a Function of mtry and ntrees",
       x = "Mtry",
       y = "Kappa",
       colour = "Ntrees") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

# Model Evaluation

## Conversion of Categorical Variables to Dummies

```{r}
dummys <- neutrofilosTest %>%
  select_if(is.factor) %>%
  dummy_columns(remove_selected_columns = TRUE, remove_first_dummy = TRUE)

colnames(dummys) <- gsub("_", "", colnames(dummys))

neutrofilosTest.dummy <- neutrofilosTest %>% 
  dplyr::select(-is.factor) %>%
  cbind(dummys)
if("Condition" %in% names(neutrofilosTest)) {
    neutrofilosTest.dummy$Condition <- factor(neutrofilosTest$Condition, levels = unique(neutrofilosTest$Condition), labels = unique(neutrofilosTest$Condition))
}
# Eliminar la columna 'Conditiontreated' de neutrofilosTest.dummy
neutrofilosTest.dummy <- neutrofilosTest.dummy[ , !names(neutrofilosTest.dummy) %in% "Conditiontreated"]
```

### Prediction with the Random Forest Model

```{r}
predict_rf <- predict(rf.cv.neutrofilos$`1500 trees`$finalModel, newdata = neutrofilosTest.dummy)
```

```{r}
matrix_rf = confusionMatrix(predict_rf, neutrofilosTest.dummy$Condition, positive = "treated")
fourfoldplot(matrix_rf$table, color = c("#CC6666", "#99CC99"), conf.level = 0, margin = 1, main = "Matriz de Confusión")
```

# Analysis of the Importance of Variables in the Random Forest Model

```{r}
importance_data <- rf.cv.neutrofilos$`1500 trees`$finalModel$importance %>%
  as.data.frame() %>%
  rownames_to_column(var = "Variable") %>%
  arrange(MeanDecreaseGini) %>%  # Ensure descending order for importance
  mutate(
    Variable = factor(Variable, levels = Variable),
    Color = ifelse(seq_along(Variable) >= 54, "red", "blue")  # Color the top 6 variables in red
  )

# Create a beeswarm plot to visualize the importance data
ggplot(importance_data, aes(x = Variable, y = MeanDecreaseGini, color = Color)) +
  geom_quasirandom(groupOnX = TRUE, size = 4, alpha = 0.6) +
  coord_flip() +
  labs(title = "Importance of the variables in the Random Forest Model",
       subtitle = "Measured by Mean Decrease Gini",
       y = "Mean Decrease Gini",
       x = "") +
  theme_minimal() +
  scale_color_identity()  # Use the colors specified in the data frame

```

