---
title: "Statistical Analysis"
subtitle: "Master in Bioinformatics, University of Murcia"
author: "Joaquín Cantón Sandoval (joaquin.canton@um.es)"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    highlight: kate
    number_sections: true
    theme: spacelab
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

# Statistical Descriptive Analysis

We start by calculating the differences in means and standard deviations for each track parameter based on whether the condition is untreated or treated.

```{r}
library(dplyr)
library(car)
library(ggplot2)
library(gridExtra)
```

```{r}
neutrophils = readRDS("neutrofilos_unique.RDS")
```

## Filter track parameters

```{r}
# Find parameters related to tracks
track_parameters <- grep("Track", names(neutrophils), value = TRUE)

to_remove <- c("TrackID", "TrackID_sd", "Track.Position.Z.Start",
              "Track.Position.Z.Mean",
                "Track.Position.Y.Start",
                "Track.Position.Y.Mean",
                "Track.Position.X.Start",
                "Track.Position.X.Mean",
               "TrackImage.CH2",
               "TrackImage.CH1", "TrackImage", "Track.Center.of.Image.Mass.Z.CH2", "Track.Center.of.Image.Mass.Z.CH1",
               "Track.Center.of.Image.Mass.Y.CH1", "Track.Center.of.Image.Mass.Y.CH2", "Track.Center.of.Image.Mass.X.CH1",
               "Track.Center.of.Image.Mass.X.CH2", "Track.Center.of.Homogeneous.Mass.Z", "Track.Center.of.Homogeneous.Mass.Y",
               "Track.Center.of.Homogeneous.Mass.X")

# Actualizar la lista de parámetros de track eliminando los no deseados
track_parameters <- setdiff(track_parameters, to_remove)

# Ver la lista actualizada
head(track_parameters)
```

## Calculate differences and SD between parameters depending on condition

```{r}
summary_stats <- neutrophils %>%
  dplyr::select(all_of(c(track_parameters, "Condition"))) %>%
  group_by(Condition) %>%
  summarise(across(all_of(track_parameters), list(mean = ~mean(.x, na.rm = TRUE), sd = ~sd(.x, na.rm = TRUE))),
            .groups = 'drop') 
```


## Levene's Test for Equality of Variances

```{r}
# Apply Levene's test to evaluate the homogeneity of the variances for each parameter
levene_results <- lapply(neutrophils[track_parameters], function(x) {
  leveneTest(x ~ neutrophils$Condition, center = median)
})

# Initialize vectors to store parameter names based on their significance
significant_parameters <- vector("character")
non_significant_parameters <- vector("character")

# Classify the parameters based on the p-value of the Levene test
for (param_name in names(levene_results)) {
  # Asegurar que estamos extrayendo correctamente el p-valor
  p_value <- levene_results[[param_name]]$`Pr(>F)`[1]
  
  # Verificar si p_value es NA antes de la comparación
  if (!is.na(p_value) && p_value <= 0.05) {
    significant_parameters <- c(significant_parameters, param_name)
  } else {
    non_significant_parameters <- c(non_significant_parameters, param_name)
  }
}
```

### Visualize variances

```{r}
# Gráfico para la variable significativa
p1 <- ggplot(neutrophils, aes(x = Condition, y = Track.Ellipsoid.Axis.C.Y.Mean)) +
  geom_boxplot(fill = "tomato", colour = "black") +
  labs(title = "Boxplot of Track.Ellipsoid.Axis.C.Y.Mean by Condition",
       x = "Condition",
       y = "Track.Ellipsoid.Axis.C.Y.Mean") +
  theme_minimal()

# Gráfico para la variable no significativa
p2 <- ggplot(neutrophils, aes(x = Condition, y = Track.Ellipsoid.Axis.C.Z.Mean)) +
  geom_boxplot(fill = "dodgerblue", colour = "black") +
  labs(title = "Boxplot of Track.Ellipsoid.Axis.C.Z.Mean by Condition",
       x = "Condition",
       y = "Track.Ellipsoid.Axis.C.Z.Mean") +
  theme_minimal()

# Combinar los dos gráficos en una cuadrícula
grid.arrange(p1, p2, nrow = 1)
```

## Mean Comparison

### Welch's T-test for Unequal Variances

For parameters that show significant differences in variances, we use Welch's t-test, which does not assume equal variances.

```{r}
# Apply Welch's t test for parameters with significantly different variances
welch_t_results <- lapply(neutrophils[significant_parameters], function(x) {
  t.test(x ~ neutrophils$Condition, var.equal = FALSE)
})
```

### Student t test for parameters with homogeneous variances

```{r}
# Apply the Student t test for the analysis for parameters with homogeneous variances
standard_t_results <- lapply(neutrophils[non_significant_parameters], function(x) {
  t.test(x ~ neutrophils$Condition, var.equal = TRUE)
})
```

## Median Comparison

### Mann-Whitney U Test for Parameters with Non-Homogeneous Variances

```{r}
# Apply the Mann-Whitney U test for parameters with significantly different variances
mann_whitney_results <- lapply(neutrophils[significant_parameters], function(x) {
  wilcox.test(x ~ neutrophils$Condition, alternative = "two.sided")
})

```

### Wilcoxon Test for Parameters with Homogeneous Variances

```{r}
# Apply the Wilcoxon test for parameters with non-significantly different variances
wilcoxon_results <- lapply(neutrophils[non_significant_parameters], function(x) {
  wilcox.test(x ~ neutrophils$Condition, alternative = "two.sided", paired = FALSE)
})
```


### Representation of mean and median differences

```{r}
plot_variance_differences <- function(data, parameters, group_column) {
  plots <- list()
  for (param in parameters) {
    p <- ggplot(data, aes_string(x = group_column, y = param, fill = group_column)) +
      geom_violin() +
      labs(title = paste("Violin Plot of", param), y = param, x = group_column) +
      theme_minimal() +
      theme(legend.position = "none") +
      stat_summary(fun = mean, geom = "point", shape = 20, size = 5, color = "lightblue", aes(group = group_column)) +
      stat_summary(fun = mean, geom = "line", linetype = "dashed", size = 2, color = "lightblue", aes(group = group_column)) +
      stat_summary(fun = median, geom = "point", shape = 23, size = 5, color = "orange", fill = "orange", aes(group = group_column)) +
      stat_summary(fun = median, geom = "line", linetype = "dashed", size = 2, color = "orange", aes(group = group_column)) +
      geom_errorbar(stat = "summary", fun.data = function(x) {
        qs <- quantile(x, probs = c(0.25, 0.75))
        list(ymin = qs[1], ymax = qs[2])
      }, width = 0.2, color = "green")  # Ajustado para cuartiles

    plots[[param]] <- p
  }
  return(plots)
}
```


```{r}
# Extract p-values ​​and associate them with variable names
p_values_wilcox <- sapply(wilcoxon_results, function(x) x$p.value)
names(p_values_wilcox) <- names(wilcoxon_results)

# Sort p-values ​​from smallest to largest
sorted_p_values_wilcox <- sort(p_values_wilcox)

# Select the first five variables with the lowest p-values
top_five_variables_wilcox <- names(sorted_p_values_wilcox)[1:5]

# Generate the plots using the plot_variance_differences function
plots <- plot_variance_differences(neutrophils, top_five_variables_wilcox, "Condition")

# Use grid.arrange to display all graphs together
plot_grid <- do.call(grid.arrange, c(plots, ncol = 2))

```

### Create a summary table of the statistical analysis

```{r}
# Function to assign significance asterisks to p-values
significance_asterisks <- function(p) {
  if (p <= 0.0001) {
    return("****")
  } else if (p <= 0.001) {
    return("***")
  } else if (p <= 0.01) {
    return("**")
  } else if (p <= 0.05) {
    return("*")
  } else {
    return("")
  }
}
# Create a dataframe with variable names, p-values ​​and asterisks for the t-test
remaining_table_t_test <- data.frame(
  Variable = names(sorted_p_values_wilcox),
  P_Value = sorted_p_values_wilcox,
  Significance = sapply(sorted_p_values_wilcox, significance_asterisks),
  stringsAsFactors = FALSE
)

# Mostrar la tabla
head(remaining_table_t_test)
```

